---
title: "Read/Plot/Query/Validate"
---

```{r setup, include=FALSE}
source("R/setup_rgl.R")
```

## Relevant Resources

-   [Code](https://github.com/liamirwin/LPS_lidRtutorial/blob/master/R/01_read.R)

-   [lidRbook section](https://r-lidar.github.io/lidRbook/io.html)

## Overview

Welcome to this lidar processing tutorial using R and the `lidR` package! In this tutorial, you will learn how to *read*, *visualize*, *query*, and *validate* lidar data. We'll explore basic information about a lidar file including the *header* and *tabular data,* as well as *visualize* point clouds using different colour schemes based on attributes. We'll use the `select` argument in `readLAS()` to load specific attributes and the `filter` argument to only load points of interest or apply transformations on-the-fly.

Let's get started with processing lidar data efficiently using `lidR` and R! Happy learning!

## Environment

We start by loading the necessary packages, clearing our current environment, and specifying that some warnings be turned off to make our outputs clearer. We will do this for each section in the tutorial.

```{r clear_warnings, warnings = FALSE, message = FALSE}
# Clear environment
rm(list = ls(globalenv()))

# Load package
library(lidR)
```

## Basic Usage

### Load and Inspect lidar Data

Load the lidar point cloud data from a LAS file using the `readLAS()` function. The data is stored in the `las` object. We can inspect the header information and attributes of the `las` object.

```{r load_and_inspect_lidar_data}
# Load the sample point cloud
las <- readLAS(files = "data/zrh_norm.laz")

# Inspect header information
las@header

# Inspect attributes of the point cloud
las@data

# Check the file size of the loaded lidar data
format(object.size(las), "Mb")
```

### Visualizing lidar Data

We can visualize the lidar data using the `plot()` function. We have several options to control the colours in the plot, such as selecting specific attributes from the data to be mapped.

::: callout-tip
## `plot()` background colour

Set the background of plots to white using `plot(las, bg = "white")`. To keep the website code clean I've omitted this from examples.
:::

``` r
plot(las)
```

```{r visualize_lidar_data, echo = FALSE, rgl = TRUE, fig.width = 8, fig.height = 6}
#| label: fig-zrh-basic
#| fig-cap: "Point cloud coloured by the Z (elevation) attribute."
# Visualize the lidar data with a default colour palette
plot(las, bg = "white")
```

``` r
plot(las, color = "Intensity")
```

```{r visualize_lidar_data_int, echo = FALSE, rgl = TRUE, fig.width = 8, fig.height = 6}
#| label: fig-zrh-intensity
#| fig-cap: "Point cloud coloured by the Intensity attribute."
# Visualize using intensity values as colours
plot(las, color = "Intensity", bg = "white")
```

``` r
plot(las, color = "Classification")
```

```{r visualize_lidar_data_class, echo = FALSE, rgl = TRUE, fig.width = 8, fig.height = 6}
#| label: fig-zrh-classification
#| fig-cap: "Point cloud coloured by Classification. Ground is blue, vegetation is green."
# Visualize using the classification attribute as colours
plot(las, color = "Classification", bg = "white")
```

``` r
plot(las, color = "ScanAngleRank")
```

```{r visualize_lidar_data_scan, echo = FALSE, rgl = TRUE, fig.width = 8, fig.height = 6}
#| label: fig-zrh-scan-angle
#| fig-cap: "Point cloud coloured by ScanAngleRank"
# Visualize using the scan angle rank as colours
plot(las, color = "ScanAngleRank", bg = "white")
```

## Point Classification

Lidar systems and pre-processing attribute point clouds with additional attributes beyond just X, Y and Z coordinates.

For instance, points are typically assigned `Classification` values to differentiate surfaces. This is often already done by the data provider or through user applied algorithms to classify for example [noise and ground](https://rdrr.io/cran/lidR/man/classify.html "lidR classify function (not covered in this workshop)") points. The meaning of these ID values depends on the LAS file version and source.

::: callout-note
LAS Standard Point classification is now standardized by the American Society for Photogrammetry and Remote Sensing (ASPRS). For a complete list of classes, refer to the [LAS 1.4 specification (R15)](https://www.asprs.org/wp-content/uploads/2019/07/LAS_1_4_r15.pdf "LAS 1.4 specification document").
:::

Our example ALS dataset, collected in 2014 over Zürich uses an older **LAS 1.1 standard**. Its classification codes are interpreted as follows:

| Class Code | Meaning | Interpretation |
|:------------------|:--------------------------------|:------------------|
| 2 | **Ground** | The bare earth surface. |
| 3/4/5 | **Low/Medium/High Vegetation** | Vegetation under 1 meter. |
| 7 | **Noise** | Erroneous points, likely below ground or outliers above ground. |
| 12 | **Overlap Points** | Points assigned to overlapping flight lines. |

### Selecting Attributes of Interest

The `readLAS()` function allows us to select specific attributes to be loaded into memory. This is useful to reduce memory requirements when working with large lidar datasets.

```{r select_attributes}
# Load only the xyz coordinates (X, Y, Z) and ignore other attributes
las <- readLAS(files = "data/zrh_norm.laz", select = "xyz")
# Inspect the loaded attributes
las@data

# Check the memory size after loading only the selected attributes
format(object.size(las), "Mb")
```

### Filtering points on-the-fly (efficient)

We can load a subset of the lidar points based on certain criteria using the `filter` argument in directly in `readLAS()`.

```{r filter_points}
# Load only the first return points
las <- readLAS(files = "data/zrh_norm.laz", filter = "-keep_first")

# Inspect the loaded points
las

# Check the memory size after loading only the filtered points
format(object.size(las), "Mb")
```

``` r
plot(las)
```

```{r visualize_lidar_data_filter, echo = FALSE, rgl = TRUE, fig.width = 8, fig.height = 6}
#| label: fig-zrh-first-returns
#| fig-cap: "Point cloud showing only first returns, filtered on load with `-keep_first`."
# Visualize the filtered points
plot(las, bg = "white")
```


The `filter` argument in `readLAS()` can be used to apply transformations on-the-fly during loading (memory efficient). This can be useful for tasks such as filtering specific classifications.

```{r on_the_fly_transform}
# Load another version of the point cloud and visualize with an applied filter
las <- readLAS(files = "data/zrh_class.laz", filter = "-keep_class 2")
```

``` r
plot(las)
```

```{r visualize_lidar_data_filter_class, echo = FALSE, rgl = TRUE, fig.width = 8, fig.height = 6}
#| label: fig-zrh-ground-filter
#| fig-cap: "Point cloud showing only ground points, filtered on readLAS with `-keep_class 2`."
plot(las, bg = "white")
```

### Filtering Points using `filter_poi()`

An alternative method for filtering points is using the `filter_poi()` function, which allows filtering based on attributes of points after the pointcloud is loaded.

```{r filter_poi_example}
# Filter points with Classification == 2
class_2 <- filter_poi(las = las, Classification == 2L)

# Combine queries to filter points with Classification 2 and ReturnNumber == 1
first_returns <- filter_poi(las = las, Classification == 2L & ReturnNumber == 1L)
```

``` r
plot(class_2)
```

```{r visualize-filter-poi-ground, echo = FALSE, rgl = TRUE, fig.width = 8, fig.height = 6}
#| label: fig-filter-poi-ground
#| fig-cap: "Ground points filtered from the LAS object using `filter_poi()`."
plot(class_2, bg = "white")
```

``` r
plot(first_returns)
```

```{r visualize-first_ground, echo = FALSE, rgl = TRUE, fig.width = 8, fig.height = 6}
#| label: fig-filter-poi-first-ground
#| fig-cap: "First-return ground points filtered using `filter_poi()`."
plot(first_returns, bg = "white")
```

## Exercises and Questions

Using:

``` r
las <- readLAS(files = "data/zrh_norm.laz")
```

#### E1.

Using the `plot()` function plot the point cloud with a different attribute that has not been done yet

Try adding axis = TRUE, legend = TRUE to your plot argument `plot(las, axis = TRUE, legend = TRUE)`

#### E2.

Create a filtered las object of returns that have an Intensity greater that 50, and plot it.

#### E3.

Read in the las file with only xyz and intensity only. Hint go to the lidRbook section to find out how to do this

## Conclusion

This concludes our tutorial on the basic usage of the `lidR` package in R for processing and analyzing lidar data. We covered loading lidar data, inspecting and visualizing the data, selecting specific attributes, and filtering points of interest.
